Left-linear rules in adhesive categories

### 2024-02-12 icalp/fusione.tex
minori

% full includes some additional material
\newif\iffull
\fulltrue
%\fullfalse

@@

% inversions
\newcommand{\inv}[1]{\ensuremath{inv}({#1})}

@@

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of distributed systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that result 
in changes to the system state. The DPO (double-pushout) approach~\cite{CMREHL:AAGT} 
is nowadays considered the standard one for these structures, thanks to its locality 
(rule application has a local effect on a state, differently e.g. from the SqPO 
approach~\cite{CorradiniHHK06}) and most importantly to its flexibility: the rules allow to 
specify that,  in a rewriting step, some entities in the current state are removed, 
some others are needed for the rewriting step to occur but remain unaltered and 
some new entities can be generated. Concerning concurrency, the chosen notion 
is shift equivalence, and independence is formally captured by the so-called 
Church-Rosser theorem of DPO rewriting~\cite[Section~xxx]{CMREHL:AAGT}.

A noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to devise 
the foundations of a rewriting theory that can be subsequently instantiated 
to the context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting. In fact, the framework is
so general that it allowed to recast in the DPO approach many additional formalisms,
ranging from toposes~\cite{johnstone2007quasitoposes} to string diagrams~\cite{bonchi2022string}.

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required to only consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagrams mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

@@

DA RIUSARE\\
The characterisation of switch equivalence based on consistent permutations is the key to
derive some simple but fundamental properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{confluence of switching}: when transforming derivation
  sequences into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;

In particular, the two properties allows one to obtain a \emph{canonical form} for switching sequences: whenever two derivation sequences are switch equivalence via some permutation $\sigma$, then each one can be obtained from the other by applying switchings to pairs of rule applications which $\sigma$ reverses (the so-called inversions).

@@

\noindent
\parbox{8cm}{
  \begin{definition}
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear DPO rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We
    will say that $\der{D}$ are $\der{D}'$ are \emph{abstraction
      equivalent} if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R\ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{H_i}}}$}

@@

@@

\begin{definition}[Switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear DPO rewriting system.  Let also
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ we write
  $\der{D} \shift{\transp{i}} \der{E}$. Then \emph{switch
    equivalence} is the least equivalence on derivation such that if
  $\der{D} \shift{\transp{i}} \der{E}$ then
  $\der{D} \equiv^s \der{E}$.
\end{definition}

@@

\begin{definition}[Good independence pair]
  \label{def:filler}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also $(i_0, i_1)$ be an independence pair between two direct
  derivations $\dder{D}_0$, $\dder{D}_1$
  as in the solid part of the diagram below
  \[
    \xymatrix@C=18pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} \ar@{->}@/^.20cm/[ddrr]|(.32)\hole|(.57)\hole_(.7){u_0} & R_0
      \ar@/^.35cm/[drrr]|(.3)\hole^(.2){i_0} \ar[dr]|(.3)\hole_{h_0}
      && L_1 \ar@/_.35cm/[dlll]_(.2){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
      \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} \ar@{->}@/_.20cm/[ddll]|(.32)\hole|(.57)\hole^(.7){u_1} && R_1 \ar[d]^{h_1} \\
      %
      G_0 &&
      \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_(.3){g_0}&& G_1 &&
      \ar@{>->}[ll]^(.3){f_1} D_1 \ar[rr]_{g_1}&& G_2\\
      %
      & & & & P \ar@/^.3cm/@{>.>}[ull]^{p_0} \ar@/_.3cm/@{.>}[urr]_{p_1}
    }
  \]

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{0,1\}$) and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ into the pullback.\todo{P: tolto ``given by \Cref{prop:tec}'', penso non sia bene avere un riferimento all'appendice qui}
  % given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{good independence pair} if
  the first two squares depicted  
  below are pushouts and if the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists\todo{P: forse si possono togliere i quadrati sotto e dire a parole?}
  \[
    \xymatrix{
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
       &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
       & P \ar@{.>}[r]_{q_1} & Q_1
    }
  \]
\end{definition}

@@

\begin{restatable}{proposition}{propcoswitch}\label{prop:coswitch}
	consistente tra cose equivalenti
\end{restatable}	
[Proof in \Cref{propcoswitch-proof}]

\begin{restatable}{lemma}{lemIde}\label{lem:ide}
	Let $\der{D}$ and $\der{D}'$ be two derivations in a linear DPO-rewriting system $(\X, \R)$ with the same length $l$. If $\id{[0,l]}:[0,l-1]\to [0,l-1]$ is a consistent permutation, then $\der{D}$ and $\der{D}'$ are abstraction equivalent.  
\end{restatable}
[Proof in \Cref{lemide-proof}]

@@

\begin{restatable}{theorem}{thmcoswitch}
Let $\X$ be an $\mathcal{M}$-adhesive category and 
$\der{D}$ and $\der{D}'$ be two derivations in a linear DPO-rewriting system $(\X, \R)$. 
Then the following are equivalent:
\begin{enumerate}
	\item $\der{D}$ and $\der{D}'$ are switch equivalent;
	\item there exists a consistent permutation $\sigma\colon [0,\lgh(D)-1]\to [0,\lgh(D)-1]$ between them.
\end{enumerate}
\end{restatable}
[Proof in \Cref{thmswitch-proof}]

@@

%The characterisation of switch equivalence based on consistent permutations is the key to
%derive some simple but fundamental properties for linear rewriting systems, namely
%\begin{enumerate}
%\item \emph{confluence of switching}: when transforming derivation
%  sequences into equivalent ones by switching independent steps, the
%  result not depends on the order in which switchings are applied, but
%  only on the resulting permutation;
  %
%\item \emph{globality of independence}: if two independent steps are
%  moved forward or backward due to the application of switchings to a
%  derivation, they remain independent.
%\end{enumerate}
%
%
%In particular, the two properties allows one to obtain a \emph{canonical form} for switching sequences: whenever two derivation sequences are switch equivalence via some permutation $\sigma$, then each one can be obtained from the other by applying switchings to pairs of rule applications which $\sigma$ reverses (the so-called inversions).\todo{P: questa prima parte da riportare in parte in intro}
%
%The failure of the characterisation of switch equivalence via
%consistent permutations for left-linear rules leads to the
%question asking whether the properties above hold.

We argued in the introduction that the characterisation of switch equivalence based on consistent permutations 
is the key to derive some simple but fundamental properties for linear rewriting systems, namely
\emph{confluence of switching} and  \emph{globality of independence}.
In this section we show that actually much of the theory can be retained: confluence of switching still holds, 
while globality of independence holds in a weak form, conceptually linked to the fact that fusions introduce a form of disjunctive causality. 
Finally, a canonical form for switching sequences can also be recovered.

We start with two lemmas dealing with derivations of length
$3$. Despite being technical, these lemma are the fundamental building blocks for proving properties (1) and (2) above.

@@

We first need some notation on permutations. A \emph{permutation} on
$\interval{n}$ is a bijection
$\sigma : \interval{n} \to \interval{n}$. We call it a
\emph{transposition} when there are $i, j \in \interval{n}$,
$i \neq j$ such that $\sigma(i)=j$, $\sigma(j) = i$ and
$\sigma(k) = k$ otherwise. Such a transposition is normally denoted as
$\transp{i}[j]$.
% Whenever $j = i+1$ we call it an \emph{adjacent transposition}.

Given a permutation $\perm$ on $\interval{n}$, the set of inversions
of $\perm$ is
$\inv{\perm} = \{ (i,j) \mid i,j \in \interval{n-1}\ \land\ i<j\
\land\ \perm(i) > \perm(j) \}$.
%
Clearly, $\perm$ is the identity iff $\inv{\perm}=0$.

We also need some notation on switching sequences. Consider derivation sequences $\der{D}_i$ for $i \in \interval[0]{n}$
related by switches
\begin{center}
  $\der{D}_0 \shift{\nu_1} \der{D}_1 \shift{\nu_2} \ldots \shift{\nu_n}
\der{D}_n$
\end{center}
It will be referred as a \emph{switching sequence}.  Given $i \leq j$,
we denote by $\nu_{i,j}$ the composition
$\nu_j \circ \nu_{i-1} \circ \ldots \nu_i$. We will say that the
switching sequence consists of inversions if for all
$i \in \interval[0]{n}$ the transposition $\nu_i$ is an inversion for
$\nu_{i,n}$.\todo{P: probabilmente questi e forse la notazione sulle
  permutazioni vanno messi quando si parla di switching, ma ora non
  ho la forza di ragionare}

@@

\subsection{Some results on $\mathcal{A}$-stable and $\mathcal{A}$-Van Kampen squares}
We will start proving some general results regarding $\mathcal{A}$-Van Kampen and $\mathcal{A}$-stable squares. Let us begin recalling some classical results about  about pullbacks and pushouts

\begin{lemma}\label{lem:popb1} \label{lem:pb1}
	Let $\X$ be a category, and consider the following diagram 	in which the right square is a pullback.
	\[\xymatrix{X \ar[d]_{a} \ar[r]^{f}& \ar[r]^{g} Y \ar[d]^{b}& Z \ar[d]^{c}\\ A \ar[r]_{h}& B \ar[r]_{k}& C}\]
	Then the whole rectangle is a pullback if and only if the left square is one.
\end{lemma}

\begin{lemma}\label{lem:po1}
	Let $\X$ be a category, and consider the following diagram 	in which the left square is a pushout.
	\[\xymatrix{X \ar[d]_{a} \ar[r]^{f}& \ar[r]^{g} Y \ar[d]^{b}& Z \ar[d]^{c}\\ A \ar[r]_{h}& B \ar[r]_{k}& C}\]
	Then the whole rectangle is a pushout if and only if the right square is one.
\end{lemma}

The following proposition establishes a key property of $\mathcal{A}$-Van Kampen squares with a mono as a side: they are not only pushouts, but also pullbacks.

\begin{proposition}\label{prop:pbpo} Let $\mathcal{A}$ be a class of arrows stable under pushouts and containing all the isomorphisms.  If $m\colon A\to C$ is mono and belongs to $\mathcal{A}$, then every $\mathcal{A}$-Van Kampen square
	\[\xymatrix{A\ar[r]^{g} \ar[d]_{m} & B \ar[d]^{n} \\ C \ar[r]_{f}  & D}\]
	is also a pullback square and $n$ is a monomorphism.
\end{proposition}

\begin{proof}
	We can start considering the cube below and noticing that $n$, being the pushout of $m\in \mathcal{M}$, belongs to $\mathcal{A}$ too.
	\[\xymatrix@C=13pt@R=13pt{&A\ar[dd]|\hole_(.65){\id{A}}\ar[rr]^{g} \ar[dl]_{\id{A}} && B \ar[dd]^{\id{B}} \ar[dl]_{\id{B}} \\ A  \ar[dd]_{m}\ar[rr]^(.65){g} & & B \ar[dd]_(.3){n}\\&A\ar[rr]|\hole^(.65){g} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{f} & & D}\]
	By construction the top face of the cube is a pushout and the back one a pullback. The left face is a pullback because $m$ is mono. Thus the $\mathcal{A}$-Van Kampen property yields that the front and the right faces are pullbacks.
\end{proof}

The previous proposition, now, allows us to establish the following result.
\begin{lemma}\label{lem:varie}Let $\mathcal{A}$ be a class of arrows stable under pullbacks, pushouts and containing all isomorphisms.  Suppose that, the left square below is $\mathcal{A}$-Van Kampen, while the vertical faces in the right cube are pullbacks.
	\[\xymatrix@C=10pt@R=10pt{&&&&&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{g'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ A \ar[dd]_{m}\ar[rr]^{g}&&B\ar[dd]^{n}&&C'  \ar[dd]_{c}\ar[rr]^(.7){f'} & & D' \ar[dd]_(.3){d}\\&&&&&A\ar[rr]|\hole^(.65){g} \ar[dl]^{m} && B \ar[dl]^{n} \\C\ar[rr]_{f} &&D&&C \ar[rr]_{f} & & D}\]
	Supppose, moreover that $m\colon A\to C$ and $d\colon D'\to D$ are mono and that $d$ belongs to $\mathcal{A}$. Then $d\leq n$ if and only if $c \leq m$.
\end{lemma}

\begin{remark}
	Recall that, given two monos $m:M\to X$ and $n:N\to X$ with the same codomain, $m\leq n$ means that there exists a, necessarily unique and necessarily mono, $k:M\to N$ fitting in the triangle below:
	\[\xymatrix{M\ar[rr]^{k}  \ar[dr]_{m}&& N \ar[dl]^{n}\\ & X}\]
	Notice, moreover, that  if $m\leq n$ and $n\leq m$, then the arrow $k:M\to N$ is an isomorphism.
\end{remark}

\begin{remark}
	Notice that, since $d$ is a mono and the vertical faces are pullbacks, then $a, b$ and $c$ are monomorphisms too. Moreover, $n$ is mono by \Cref{prop:pbpo}, so that even $m'$ and $n'$ are monos.
\end{remark}

>>>>

% full includes some additional material
\newif\iffull
%\fulltrue
\fullfalse

@@

% inversions
\newcommand{\inv}[1]{\mathsf{inv}({#1})}

@@

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that result 
in changes to the system state. The DPO (double-pushout) approach~\cite{CMREHL:AAGT} 
is nowadays considered the standard one for these structures, thanks to its locality 
(rule application has a local effect on a state, differently e.g. from the SqPO 
approach~\cite{CorradiniHHK06}) and most importantly to its flexibility: the rules allow to 
specify that,  in a rewriting step, some entities in the current state are removed, 
some others are needed for the rewriting step to occur but remain unaltered and 
some new entities can be generated. Concerning concurrency, the chosen notion 
is shift equivalence, and independence is formally captured by the so-called 
Church-Rosser theorem of DPO rewriting~\cite[Section~xxx]{CMREHL:AAGT}.

A noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to devise 
the foundations of a rewriting theory that can be subsequently instantiated 
to the context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting. In fact, the framework is
so general that it allowed to recast in the DPO approach many additional formalisms,
ranging from xxx to string diagrams~\cite{bonchi2022string}.

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required only to consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagram mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

@@

@@

\noindent
\parbox{8cm}{
  \begin{definition}
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear DPO rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We
    will say that $\der{D}$ are $\der{D}'$ are \emph{abstraction
      equivalent}, and use the symbol $\der{D}\equiv_a \der{D}'$, if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R\ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{H_i}}}$}

@@

\todo{Se inseriamo le figure in fusione.tex si risparmia dello spazio}
  \begin{figure}
    \input{Figure/rules.tex}

@@

We first need to recall some concept and notations regarding permutations.

\begin{definition}
	A \emph{permutation} on
	$\interval[0]{n}$ is a bijection
	$\sigma : \interval[0]{n} \to \interval[0]{n}$. We call it a
	\emph{transposition} when there are $i, j \in \interval{n}$,
	$i \neq j$ such that $\sigma(i)=j$, $\sigma(j) = i$ and
	$\sigma(k) = k$ otherwise. Such a transposition is normally denoted as
	$\transp{i}[j]$.

Given a permutation $\perm$ on $\interval[0]{n}$,  an \emph{inversion} for $\sigma$ is a pair $(i,j)$ such that $i<j$ and $\sigma(j)< \sigma(i)$. The set of inversions
of $\perm$ will be denoteb by $\inv{\sigma}$.
\end{definition}

\begin{remark}
	Clearly, a permutation $\perm$ is the identity if and only if $\inv{\perm}=\emptyset$.
\end{remark}

We also need to introduce the notion of \emph{switching sequences}.

\begin{definition}
	Consider derivations $\der{D}_i$ for $i \in \interval[0]{n}$
	related by switches
	\begin{center}
		$\der{D}_0 \shift{\nu_1} \der{D}_1 \shift{\nu_2} \ldots \shift{\nu_n}
		\der{D}_n$
	\end{center}
	It will be referred as a \emph{switching sequence}.  
\end{definition}

\begin{definition}[Switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear DPO rewriting system.  Let also
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ we write
  $\der{D} \shift{\transp{i}} \der{E}$.

  A \emph{switching sequence} is  a sequence $\{\der{D}_{k}\}_{i=0}^m$ of derivations such that 
  \[\der{D}_0 \shift{\transp{i_1}} \der{D}_1 \shift{\transp{i_2}} \ldots \shift{\transp{i_m}}
  \der{D}_m\]
 In such case, we will often denote by $\nu_{k}$ the transposition $\transp{i_k}$. Given $k \leq h$,
 we denote by $\nu_{k,h}$ the composition
 $\nu_h \circ \nu_{h-1} \circ \ldots \nu_k$. We will say that the
 switching sequence consists of inversions if for all
 $k \in \interval[0]{m}$ the transposition $\nu_k$ is an inversion for
 $\nu_{1,m}$.

  We will say that two derivations $\der{D}, \der{E}:G\Mapsto H$ are \emph{switch
    equivalent} if there is a switching sequence, possibly empty, $\{\der{D}_{k}\}_{i=0}^m$ such that
      \[\der{D}\equiv_a \der{D}_0 \shift{\transp{i_1}} \der{D}_1 \shift{\transp{i_2}} \ldots \shift{\transp{i_m}}
    \der{D}_m\equiv _a \der{E}\]
\end{definition}

\begin{remark}\label{rem:abst}
	The possibility of an empty switching sequence assures that two abstraction equivalent derivations are switch equivalent.
\end{remark}

@@

\begin{definition}[Good independence pair]
  \label{def:filler}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also $(i_0, i_1)$ be an independence pair between two direct
  derivations $\dder{D}_0$, $\dder{D}_1$
  as in the solid part of the diagram below
  \[
    \xymatrix@C=18pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} \ar@{.>}@/^.20cm/[ddrr]|(.32)\hole|(.57)\hole_(.7){u_0} & R_0
      \ar@/^.35cm/[drrr]|(.3)\hole^(.2){i_0} \ar[dr]|(.3)\hole_{h_0}
      && L_1 \ar@/_.35cm/[dlll]_(.2){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
      \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} \ar@{.>}@/_.20cm/[ddll]|(.32)\hole|(.57)\hole^(.7){u_1} && R_1 \ar[d]^{h_1} \\
      %
      G_0 &&
      \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_(.3){g_0}&& G_1 &&
      \ar@{>->}[ll]^(.3){f_1} D_1 \ar[rr]_{g_1}&& G_2\\
      %
      & & & & P \ar@/^.3cm/@{>.>}[ull]^{p_0} \ar@/_.3cm/@{.>}[urr]_{p_1}
    }
  \]

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{0,1\}$) and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ into the pullback (see \Cref{prop:tec} for details). 
  % given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{good independence pair} if
  the first two squares depicted  
  below are pushouts and if the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists.
  \[
    \xymatrix{
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
       &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
       & P \ar@{.>}[r]_{q_1} & Q_1
    }
  \]
\end{definition}

@@

\begin{remark}\label{rem:abs}
	Let $\der{D}$ and $(\der{E}, \alpha', \omega')$ be two abstraction equivalent derivations of length $n$. Then the identity permutation $\id{[0,n-1]}\colon [0,n-1]\to [0,n-1]$ is a consistent permutation between them. In such a case we can take as $\xi_{\id{[0,n-1]}}\colon \tpro{D}\to \lpro \der{D}'\rpro$ simply the isomorphism induced by any abstraction equivalence.
\end{remark}

\begin{remark}\label{rem:compo}
	It is immediate to see that consistent permutations $\sigma$ and $\tau$ compose: as mediating isomorphism it is enough to take the composite of $\xi_{\sigma}$ and $\xi_\tau$. Similarly, $\xi^{-1}_{\sigma}$ witnesses that $\sigma^{-1}$ is a consistent permutation whenever $\sigma$ is so.
\end{remark}
In a linear rewriting system the converse also holds.

\begin{restatable}{lemma}{lemIde}\label{lem:ide}
	Let $\der{D}$ and $\der{D}'$ be two derivations in a linear DPO-rewriting system $(\X, \R)$ with the same length $l$. If $\id{[0,l-1]}:[0,l-1]\to [0,l-1]$ is a consistent permutation, then $\der{D}$ and $\der{D}'$ are abstraction equivalent.  
\end{restatable}
[Proof in \Cref{lemide-proof}]

@@

\begin{restatable}{proposition}{propcoswitch}\label{prop:coswitch}
Let $\der{D}$ and $\der{E}$ be two switch equivalent derivations, then there exists a consistent permutation between them.
\end{restatable}	
[Proof in \Cref{propcoswitch-proof}]

In the linear case, the existence of a consistent permutation it is enough to characterize switch equivalence. \todo{FORSE QUA METTEREI DEI POINTER A QUALCHE LAVORO PASSATO?}

\begin{restatable}{theorem}{thmcoswitch}
Let $\X$ be an $\mathcal{M}$-adhesive category and 
$\der{D}$ and $\der{D}'$ be two derivations in a linear DPO-rewriting system $(\X, \R)$. 
Then the following are equivalent:
\begin{enumerate}
	\item $\der{D}$ and $\der{D}'$ are switch equivalent;
	\item there exists a consistent permutation $\sigma\colon [0,\lgh(\der{D})-1]\to [0,\lgh(\der{D})-1]$ between them.
\end{enumerate}
\end{restatable}
[Proof in \Cref{thmswitch-proof}]

@@

The characterisation of switch equivalence based on consistent permutations is the key to
derive some simple but fundamental properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{confluence of switching}: when transforming derivation
  sequences into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;

In particular, the two properties allows one to obtain a \emph{canonical form} for switching sequences: whenever two derivation sequences are switch equivalence via some permutation $\sigma$, then each one can be obtained from the other by applying switchings to pairs of rule applications which $\sigma$ reverses (the so-called inversions).\todo{P: questa prima parte da riportare in parte in intro}

In this section we show that actually much of the theory can be retained: confluence of switching still holds, while globality of independence holds in a weak form, conceptually linked to the fact that fusions introduce a form of disjunctive causality. Finally, a canonical form for switching sequences can also be recovered.

We start by proving two lemmas dealing with derivations of length
$3$. Despite being technical, these lemma are the fundamental building blocks for showing properties (1) and (2) above.

@@

@@

\subsection{Some results on $\mathcal{A}$-stable and $\mathcal{A}$-Van Kampen squares}
We will start proving some general results regarding $\mathcal{A}$-Van Kampen and $\mathcal{A}$-stable squares. Let us begin recalling some classical results about  about pullbacks and pushouts\cite{mac2013categories}.

\noindent 
\parbox{10.5cm}{\begin{lemma}\label{lem:popb1} \label{lem:pb1} \label{lem:po1}
Let $\X$ be a category, and consider the diagram  aside, then the following are true:
	\begin{enumerate}
		\item if the right square is a pullback, then the whole rectangle is a pullback if and only if the left square is one;
		\item if the left square is a pushout, 	then the whole rectangle is a pushout if and only if the right square is one.
	\end{enumerate}
\end{lemma}}
\parbox{4cm}{$\xymatrix{X \ar[d]_{a} \ar[r]^{f}& \ar[r]^{g} Y \ar[d]^{b}& Z \ar[d]^{c}\\ A \ar[r]_{h}& B \ar[r]_{k}& C}$}\\

The following proposition establishes a key property of $\mathcal{A}$-Van Kampen squares with a mono as a side: they are not only pushouts, but also pullbacks \cite{ehrig2004adhesive,behr2022fundamentals,lack2005adhesive}.

\noindent 
\parbox{11.5cm}{\begin{proposition}\label{prop:pbpo} Let $\mathcal{A}$ be a class of arrows stable under pushouts and containing all the isomorphisms.  If $m\colon A\to C$ is mono and belongs to $\mathcal{A}$, then every $\mathcal{A}$-Van Kampen square
	is also a pullback square and $n$ is a monomorphism.
\end{proposition}}
\parbox{2cm}{\vspace{1.5ex}$\xymatrix{A\ar[r]^{g} \ar[d]_{m} & B \ar[d]^{n} \\ C \ar[r]_{f}  & D}$}

\full{ 
\begin{proof}
	We can start considering the cube below and noticing that $n$, being the pushout of $m\in \mathcal{M}$, belongs to $\mathcal{A}$ too.
	\[\xymatrix@C=13pt@R=13pt{&A\ar[dd]|\hole_(.65){\id{A}}\ar[rr]^{g} \ar[dl]_{\id{A}} && B \ar[dd]^{\id{B}} \ar[dl]_{\id{B}} \\ A  \ar[dd]_{m}\ar[rr]^(.65){g} & & B \ar[dd]_(.3){n}\\&A\ar[rr]|\hole^(.65){g} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{f} & & D}\]
	By construction the top face of the cube is a pushout and the back one a pullback. The left face is a pullback because $m$ is mono. Thus the $\mathcal{A}$-Van Kampen property yields that the front and the right faces are pullbacks.
\end{proof}}

The previous proposition, now, allows us to establish the following result.

\noindent \parbox{7.4cm}{
\begin{lemma}\label{lem:varie}Let $\mathcal{A}$ be a class of arrows stable under pullbacks, pushouts and containing all isomorphisms.  Suppose that, the left square below is $\mathcal{A}$-Van Kampen, while the vertical faces in the right cube are pullbacks. 	Supppose, moreover that $m\colon A\to C$ and $d\colon D'\to D$ are mono and that $d$ belongs to $\mathcal{A}$. Then $d\leq n$ if and only if $c \leq m$.
\end{lemma}}
	\parbox{6cm}{$\xymatrix@C=10pt@R=10pt{&&&&&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{g'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ A \ar[dd]_{m}\ar[rr]^{g}&&B\ar[dd]^{n}&&C'  \ar[dd]_{c}\ar[rr]^(.7){f'} & & D' \ar[dd]_(.3){d}\\&&&&&A\ar[rr]|\hole^(.65){g} \ar[dl]^{m} && B \ar[dl]^{n} \\C\ar[rr]_{f} &&D&&C \ar[rr]_{f} & & D}$}

\noindent 
\parbox{11cm}{\begin{remark}
	Recall that, given two monos $m:M\to X$ and $n:N\to X$ with the same codomain, $m\leq n$ means that there exists a, necessarily unique and necessarily mono, $k:M\to N$ fitting in the triangle aside.

	\hspace{15pt}Notice, moreover, that  if $m\leq n$ and $n\leq m$, then the arrow $k:M\to N$ is an isomorphism.
\end{remark}}
\parbox{4cm}{$\xymatrix@C=15pt{M\ar@{.>}[rr]^{k}  \ar[dr]_{m}&& N \ar[dl]^{n}\\ & X}$}\\
